---
title: "Radon Data Exploratory Analysis"
output: html_document
---

```{r setup, eval=TRUE, echo=TRUE, message=FALSE}
library(tidyverse)
library(rbokeh)
# devtools::install_github("hafen/trelliscopejs")
library(trelliscopejs)
```

# Problem Description

Here's the problem description from D3M:

> Radon is a carcinogen -- a naturally occurring radioactive gas whose decay products are also radioactive -- known to cause lung cancer in high concentrations and estimated to cause several thousand lung cancer deaths per year in the United States. The distribution of radon levels in U.S. homes varies greatly, with some houses having dangerously high concentrations. To identify areas of high radon exposure, the Environmental Protection Agency coordinated radon measurements in a random sample of houses throughout the country.
>
> The overall goal is to estimate the distribution of radon levels across U.S. counties, so that homeowners could make decisions about measuring or remediating the radon in their houses based on the best available knowledge of local conditions. For the purpose of this problem, we had an important predictor - whether the measurement was taken in a basement. (Radon comes from underground and can enter more easily when a house is built into the ground.) We also had an important county-level predictor - a measurement of soil uranium that was available at the county level. In this problem we'll look at Minnesota, a state that contains 85 county's in which different measurements are taken, ranging from 2 till 80 measurements per county. Build a model to make predictions of radon levels based on the county data and the presence of a basement.

# Data

There are two sources of data. One is from Andrew Gelman's [hierarchical modeling book](http://www.stat.columbia.edu/~gelman/arm/) and the other from D3M. Gelman's data has results across the United States, while the D3M data is just for Minnesota.

```{r read, message=FALSE}
# from Gelman
radon <- read_csv("data/ARM/srrs2.dat", na = c(".", ""))
radon

# from D3M
radon_mn <- read_csv("data/r_26/raw_data/radon.csv")
radon_mn

# how do these datasets differ in terms of variables?
setdiff(names(radon), names(radon_mn))
setdiff(names(radon_mn), names(radon))
```

I haven't found a data dictionary for these. The D3M schema doesn't provide any information about the variables.

Here's a summary of each of the variables:

```{r summ, eval=TRUE, echo=TRUE}
summary(radon)
```

The data from Gelman also provides geographical coordinates for the counties:

```{r geo, eval=TRUE, echo=TRUE, message=FALSE}
geo <- read_csv("data/ARM/cty.dat")
geo
```

# Exploration

## Response variable: activity

Let's look at the response variable `activity`. From the summary above, we see that it is non-negative. A quick look at its distribution:

```{r explore1, eval=TRUE, echo=TRUE}
plot(sort(radon$activity))
```

It looks very heavy-tailed so we probably want to look into a log transformation. But there are zero-valued observations. Also, are there any missing values?

```{r explore2, eval=TRUE, echo=TRUE}
length(which(is.na(radon$activity)))
```

No missing values. Let's see how many zero-valued observations there are:

```{r explore3, eval=TRUE, echo=TRUE}
length(which(radon$activity == 0))
```

Not many. What's the lowest non-zero value we observe?

```{r explore4, eval=TRUE, echo=TRUE}
min(radon$activity[radon$activity > 0])
```

Let's create a log activity variable and deal with zeros by setting them to the lowest-observed non-zero value.

```{r explore5, eval=TRUE, echo=TRUE}
radon$log_activity <- log2(ifelse(radon$activity == 0, 0.1, radon$activity))
```

Now let's look at the distribution of this variable:

```{r explore6, eval=TRUE, echo=TRUE}
plot(sort(radon$log_activity))
```

This looks better. It may even look normally distributed. Let's check:

```{r explore7, eval=TRUE, echo=TRUE}
qqnorm(radon$log_activity)
qqline(radon$log_activity)
```

We'll use this variable in our modeling.

## Other covariates

Here are some exploratory visualizations of some of the other covariates.

Some of these variables, while they look like numeric, are treated better as categorical, so we'll change their type before plotting:

```{r explore9, eval=TRUE, echo=TRUE}
radon$room <- as.character(radon$room)
radon$floor <- as.character(radon$floor)
radon$typebldg <- as.character(radon$typebldg)
```

### wave

I don't know what the `wave` variable is. It's numeric so let's look at its distribution.

```{r explore8, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_quantile(wave)
figure(data = radon) %>% ly_hist(wave)
```
The distribution has two modes...

### room

```{r explore12, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_bar(room, hover = TRUE)
```

Strange that there aren't many observed room values of 5 and 6 when there's a significant amount of 7.

### floor

```{r explore13, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_bar(floor, hover = TRUE)
```

Why is there a floor 9 but nothing between 3 and 9?

### typebldg

```{r explore14, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_bar(typebldg, hover = TRUE)
```

I don't know what this is, but 1 is the most prominent (maybe that's residential?).

### basement

```{r explore15, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_bar(basement, hover = TRUE)
```

I wonder what "0" means...

### Variable combinations

```{r explore16, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_bar(room, color = basement, hover = TRUE)
```

```{r explore17, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_bar(room, color = basement, hover = TRUE, position = "fill")
```

```{r explore18, eval=TRUE, echo=TRUE}
figure(data = radon) %>% ly_bar(typebldg, color = room, hover = TRUE, position = "fill")
```
